Weakly informative observations: not the winner

### 2020-02-04 inst/notthewinner.Rmd
total reworking

Here we consider a situation in which we make the observation that a
particular player *does not* win.  We are ignorant of the order
statistic except for the fact that (wlog) player 1 is not first
placed.

@@

* $H_1\colon p_1=0$
* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
* $H_3\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$

(observe that $H_2$ implies that someone is worse than me).  All three
hypotheses correspond to a restriction on allowable parameter space.
The first step is to define a suitable likelihood function for our
observation.  Given that $P(D|H_1)=1$, $P(D|H_2)\leq\frac{1}{2}$ and
$P(D|H_3) =\frac{n-1}{n}$, this gives us a likelihood function for our
three hypotheses.

What we will do is combine this observation with other, perhaps more
informative, observations which we will take to be (conditionally)
independent of one another.  Firstly we will consider some highly
informative order statistics:

```{r}
(H <- hyper2(pnames=paste("p",1:5,sep="")) +
    order_likelihood(c(5,2,1,4)) + 
    order_likelihood(c(2,4,1)) + 
    order_likelihood(c(4,3,1)) + 
    order_likelihood(c(3,5)))
(mH <- maxp(H))
loglik(indep(mH),H)
```

Note the low estimated strength of `p1`.

@@

```{r}
small <- 1e-3
(mH1 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH1),H)
```

and for a consistency we will use slightly different values of `small`:

```{r}
small <- 5e-4
(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH2),H)
```

```{r}
small <- 3e-4
(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH2),H)
```

Observe that decreasing the value of variable `small` corresponds to a
more restrictive constraint on the likelihood.  And as the value of
`small` decreases the likelihood at the evaluate decreases quite
substantially (the constraint is active as we can see from the first
element of the evaluate).  Usually this does not matter but in this
case we can deduce using elementary arguments that the likelihood of
$H_2$ is zero: if $p_1=0$, then $p_4=0$ from the first line, and then
$p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$ from line 4 and
finally $p+2=0$ from line 1.

Thus we can say for sure that $H_1$ has zero likelihood and we may
discount it.

@@

```{r}
n <- 5 # number of players
small <- 1e-3  # numerical necessity
(ans_constrained <-
maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
)
small <- 5e-4
(ans_constrained <-
maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
)
```

In the above, note that the value of `small` is immaterial and the
evaluate has $p_1=p_2\neq 0$.

## Hypothesis 3

```{r}
loglik(indep(equalp(H)),H)
```

The support for $H_2$ against $H_3$ is about $-6.846-(-7.455) = 1.391$.

# Additional observation

Suppose we now make another observation: person 1 did not win.  How
does this change our relative support for $H_2$ and $H_1$?

```{r}
library("partitions")
M <- perms(5)
SL <- list()
for(i in seq_len(ncol(M))){
	SL[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
}
SL <- as.suplist(SL)
1e4*like_single_list(indep(equalp(SL)),SL)
1e4*like_single_list(indep(maxp(SL)),SL)
```

Something is very wrong here because the support should be very
similar to the support applied to the order statistic without the
additional observation.

>>>>

Here we consider a situation in which we make an observation that is
only very very weakly informative.  We consider the observation
${\mathcal O}$ that a particular player *does not* win: the order
statistic is unobserved and we know nothing about it except for the
fact that (wlog) player 1 (aka "me") is not first placed.

@@

* $H_0\colon\sum p_i=1$  (that is, no restriction)
* $H_1\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
* $H_2\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$

(see how $H_1$ implies that someone is worse than me).  Observe that
$H_1$ and $H_2$ correspond to a restriction on allowable parameter
space.  The first step is to define a suitable likelihood function for
our data.

What we will do is combine ${\mathcal O}$ with other, perhaps more
informative, observations which we will take to be (conditionally)
independent of one another.  Firstly we will consider some highly
informative order statistics:

```{r}
(H <- hyper2(pnames=paste("p",1:5,sep="")) +
    order_likelihood(c(5,2,1,4)) + 
    order_likelihood(c(2,4,1)) + 
    order_likelihood(c(4,3,1)) + 
    order_likelihood(c(3,5)))
(mH0 <- maxp(H))
loglik(indep(mH0),H)
```

Above, `mH0` corresponds to $H_0$.  Note the low but nonzero estimated
strength of `p1`.

Observe in passing that we may reject $H_z\colon p_1=0$
unconditionally: we can deduce using elementary arguments that the
likelihood of $H_z$ is zero: if $p_1=0$, then $p_4=0$ from the first
line, and then $p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$
from line 4 and finally $p_2=0$ from line 1.  Thus we can say for sure
that $H_z$ is incorrect and we may safely ignore it henceforth.

@@

Hypothesis 1 is rather tricky so we will use the ersatz $H_{1'}\colon
p_1\geqslant p_4$ (we choose $p_4$ on the grounds of its low estimated
value).  Then

```{r}
n <- 5 # number of players
small <- 1e-3  # numerical necessity
(mH1 <- maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0))
loglik(indep(mH1),H)
```

(the numerical value of `small` is immaterial).

@@

```{r}
mH2 <- equalp(H)
loglik(indep(mH2),H)
```

## Support function

We can present a support function for $H_0,H_1,H_2$:

```{r,suppcalculator}
Supp <- c(H0=loglik(indep(mH0),H), H1=loglik(indep(mH1),H), H2=loglik(indep(mH2),H)
)
Supp-max(Supp)
```

# Additional uninformative observation

Suppose we now make a perfectly uninformative observation: in an
independent trial, the order statistic for the five competitors was
one of the $5!=120$ possible orders.  How does this change our
relative support for $H_0$ against $H_1$ and $H_2$?  It should make no
difference.

```{r,uninformativesupportfunction}
library("partitions")
H # as before, here for convenience
M <- perms(5)   # uninformative!
SL <- list()
for(i in seq_len(ncol(M))){
	SL[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
}
SL <- as.suplist(SL)
```

So `SL` is a compound object that gives a likelihood function for the
joint observation of the order statistics and the uninformative data a
subsequent (independent) trial was in any of the $5!=120$ possible
orders.  We can find the MLE for different observations:

```{r,mls0}
mSL0 <- maxp(SL)
mSL1 <- maxp(SL,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
mSL2 <- equalp(SL)
```

We use `like_single_list()` to calculate the likelihoods at the
different evaluates; note that this function returns likelihoods, not
supports.

```{r,supp2calc}
Supp2 <-
      log(c(
      H0 = like_single_list(indep(mSL0),SL),
      H1 = like_single_list(indep(mSL1),SL),
      H2 = like_single_list(indep(mSL2),SL)))
(Supp2 <- Supp2-max(Supp2))
```

which is identical to the previous support function.

# Additional weakly informative observation

Now we make another observation ${\mathcal O}$ that is weakly
informative: that person `1` did not win:

```{r,slacalc}
M <- perms(5)
M <- M[, M[1,]!=1]  # remove any column with "1" in first place
SLa <- list()
for(i in seq_len(ncol(M))){
	SLa[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
}
SLa <- as.suplist(SLa)
```

So `SLa` is a compound object that gives a likelihood function for the
joint observation of the order statistics and ${\mathcal O}$.  We can
find the MLE for different observations:

```{r,mlsa}
mSLa0 <- maxp(SLa)
mSLa1 <- maxp(SLa,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
mSLa2 <- equalp(SLa)
```

```{r,suppacalc}
Suppa <-
      log(c(
      H0 = like_single_list(indep(mSLa0),SLa),
      H1 = like_single_list(indep(mSLa1),SLa),
      H2 = like_single_list(indep(mSLa2),SLa)))
(Suppa <- Suppa-max(Suppa))
```

To compare:

```{r,comparesups}
Supp2-Suppa
```

Thus ${\mathcal O}$ furnishes a tiny bit of extra evidence against
$H_1$ (about $0.016$ units of support) and a little bit of extra
evidence against $H_2$ (about $0.193$ units of support).
